---
title: "Vista Teleportation Analysis: Spatial Cognition in Virtual Environments"
author: "Your Name"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    self-contained: true
    df-print: kable
---

# Introduction
```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(report)
library(broom)
library(gt)
library(ggeffects)
library(performance)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

ROUNDING_DIGITS <- 3

here::i_am("reports/analysis.qmd")
source(here::here("functions", "visualisation.R"))
source(here::here("reports", "functions", "reporting.R"))

PROCESSED_DATE <- "20260301"

df_pointing <- read.csv(here::here("temp", "processed", paste0("pointing", PROCESSED_DATE, ".csv")))
df_distance <- read.csv(here::here("temp", "processed", paste0("distance", PROCESSED_DATE, ".csv")))
df_timing <- read.csv(here::here("temp", "processed", paste0("timing", PROCESSED_DATE, ".csv")))

df_trials <- read.csv(here::here("temp", "processed", paste0("all_combined_trials", PROCESSED_DATE, ".csv")))
df_data <- read.csv(here::here("temp", "processed", paste0("all_combined", PROCESSED_DATE, ".csv")))

df_trials <- df_trials %>%
  mutate(LevelName = str_c(str_remove(LevelName, "Vista|Nonvista"), "_", LevelSize),
         Vista = factor(Vista, levels = c("Nonvista", "Vista")),
         LevelSize = factor(LevelSize, levels = c("Small", "Large")),
         is_training = (target == "OriginDoor"),
         run = gsub("VT", "run2_", participant),
         run = gsub("_.*", "", run),
         pointing_duration = as.numeric(FirstPointing_time - start_time),
         OpticFlow = factor(Movement, levels = c("Teleport", "OpticFlow"), labels = c("Without OF", "With OF")))

df_data <- df_data %>%
  mutate(run = gsub("VT", "run2_", participant),
         run = gsub("_.*", "", run),
         Vista = factor(Vista, levels = c("Nonvista", "Vista")),
         LevelSize = factor(LevelSize, levels = c("Small", "Large")),
         OpticFlow = factor(Movement, levels = c("Teleport", "OpticFlow"), labels = c("Without OF", "With OF")))

df_trials_model <- df_trials %>%
  filter(!is_training)

df_questionnaires <- df_data %>%
  select(participant, starts_with("ssq_"), starts_with("vrsq_"), starts_with("vrleq_"), Movement, Gender) %>%
  mutate(OpticFlow = factor(Movement, levels = c("Teleport", "OpticFlow"), labels = c("Without OF", "With OF"))) %>%
  distinct()

#' To ensure model stability, we set explicit treatment contrasts for all factors to be 0 1 and not mean -centred.
#' 
#' Set treatment contrasts for model stability
#' name the contrasts explicitly
contrasts(df_trials_model$OpticFlow) <- c(0, 1)
contrasts(df_trials_model$LevelSize) <- c(0, 1)
contrasts(df_trials_model$Vista) <- c(0, 1)
```

This report presents the analysis of spatial cognition in virtual environments using the Vista Teleportation paradigm. The study examines how different movement modes (with and without optic flow) and environmental configurations (vista vs. non-vista spaces, small vs. large environments) affect spatial navigation performance.

## Model presentation interpretation

We have used linear mixed-effects models (LMMs) and generalized linear mixed-effects models (GLMMs) to analyze the data where possible. Participants were included as random effects to account for individual differences. The GLMMs were used in situations where the outcome variable exhibited non-normal distributions (e.g., angle errors, response times) and in such cases we used Gamma distribution with log link function.

Some outcome measures were not suitable for mixed-effects modeling due to limited observations per participant per condition (e.g., total navigation distance and duration). In these cases, we employed generalized linear models (GLMs) with appropriate distributions (Gamma with log link or Gaussian) to analyze the data at the group level.

In all models the reference levels and intercepts are as follows:

- Optic Flow: Without Optic Flow (Teleportation)
- Level Size: Small
- Vista: Non-vista
  
Any beta coefficients represent the change in the outcome variable when moving from the reference level to the other level of the factor, holding all other variables constant. For clarity and clearer interpretation of interactions, marginal means and predicted values are also reported.

# Participant Characteristics

```{r descriptives}
#| tbl-cap: "Demographic characteristics of participants by gender"
df_data %>%
  select(participant, Age, Gender) %>%
  distinct() %>%
  group_by(Gender) %>%
  summarise(
    `Mean Age` = round(mean(Age), ROUNDING_DIGITS),
    `SD Age` = round(sd(Age), ROUNDING_DIGITS),
    `Min Age` = min(Age),
    `Max Age` = max(Age),
    N = n(),
    .groups = "drop"
  ) %>%
  gt() %>%
  tab_header(title = "Participant Demographics")
```

```{r demographics-calc}
# Calculate demographic statistics for inline reporting
n_total <- n_distinct(df_data$participant)
demo_data <- df_data %>%
  select(participant, Age, Gender) %>%
  distinct()
n_female <- sum(demo_data$Gender == "Female")
n_male <- sum(demo_data$Gender == "Male")
mean_age <- round(mean(demo_data$Age), ROUNDING_DIGITS)
sd_age <- round(sd(demo_data$Age), ROUNDING_DIGITS)
```

A total of `r n_total` participants completed the study. The sample consisted of `r n_female` females and `r n_male` males, with an average age of `r mean_age` years (SD = `r sd_age`).

# Data Quality Validation

## Training Trial Performance

Before analyzing the main experimental trials, we validated participant performance on training trials where they pointed to the origin door. This serves as a basic competency check to ensure participants understood the pointing task.

```{r validation-origin-door}
#| fig-cap: "Distribution of pointing errors during training (origin door) vs. experimental trials. Left panels show training trials, right panels show experimental trials."
df_trials %>%
  select(participant, target, is_training, target_pointed_distance_difference,
         pointed_angle_difference) %>%
  pivot_longer(cols = target_pointed_distance_difference:pointed_angle_difference,
               names_to = "metric", values_to = "value") %>%
  mutate(
    metric = case_when(
      metric == "target_pointed_distance_difference" ~ "Distance Error (m)",
      metric == "pointed_angle_difference" ~ "Angle Error (degrees)"
    ),
    is_training = ifelse(is_training, "Training", "Experimental")
  ) %>%
  ggplot(aes(x = value, fill = is_training)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  facet_wrap(~metric + is_training, scales = "free", ncol = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Pointing Accuracy: Training vs. Experimental Trials",
    x = "Error Magnitude",
    y = "Frequency"
  ) +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")
```

Participants demonstrated good accuracy on training trials, with relatively small errors in both distance and angular pointing, confirming they understood the task requirements.

## Experimental Design Validation

```{r validation-run}
#| tbl-cap: "Training trial performance by experimental run (verbal instruction variant)"
df_trials %>%
  filter(!is_training) %>%
  group_by(run) %>%
  summarize(
    `Mean Distance Error (m)` = round(mean(target_pointed_distance_difference, na.rm = TRUE), 2),
    `SD Distance Error` = round(sd(target_pointed_distance_difference, na.rm = TRUE), 2),
    `Mean Angle Error (deg)` = round(mean(abs(pointed_angle_difference), na.rm = TRUE), 2),
    `SD Angle Error` = round(sd(abs(pointed_angle_difference), na.rm = TRUE), 2),
    N = n(),
    .groups = "drop"
  ) %>%
  gt() %>%
    tab_header(title = "Training Trial Performance by Experimental Run")

aov(target_pointed_distance_difference ~ run, data = df_trials %>% filter(!is_training)) %>%
  summary()
```

```{r checking-distnace-diffence}
df_trials %>%
  filter(target_pointed_distance_difference < 1000) %>%
ggplot(., aes(target_pointed_distance_difference, fill = run)) +
  geom_histogram(position = "identity", bins = 30, alpha = 0.5) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Pointing Distance Errors by Experimental Run",
    x = "Distance Error (meters)",
    y = "Frequency",
    fill = "Experimental Run"
  ) +
  facet_wrap(~is_training) +
  scale_fill_brewer(palette = "Set1")


## make the same graph but boxplots
df_trials %>%
  filter(target_pointed_distance_difference < 1000) %>%
  ggplot(aes(x = run, y = target_pointed_distance_difference, fill = run, color = run)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~is_training) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Pointing Distance Errors by Experimental Run",
    y = "Distance Error (meters)",
    x = "Experimental Run",
    fill = "Experimental Run",
    color = "Experimental Run"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```

Performance on training trials was consistent across different experimental runs, suggesting that verbal instruction variations did not systematically bias participant performance.

# Main Analyses

## Pointing Distance Error

We first examined how accurately participants estimated distances to target objects. The distance error represents the difference between the actual distance to the target and the distance to the location they indicated.


```{r pointing-distance-histogram}
#| fig-cap: "Distribution of pointing distance errors showing potential outliers"
ggplot(aes(x = target_pointed_distance_difference), data = df_trials_model) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Pointing Distance Errors",
    x = "Distance Error (meters)",
    y = "Frequency"
  )
```

### Statistical Model

We fitted a linear mixed-effects model to examine the effects of optic flow, environment size, and vista condition on pointing distance errors. The model included the actual distance to the target as a covariate to account for the fact that larger distances might lead to larger errors.

Extreme outliers (|z| > 3) were excluded from the analysis.

Regular mixed effect model with normal distrubution has been used, as the distribution of distance errors was approximately normal around 0-mean. 

```{r lmer-pointed-distance, results='hide'}
df_distance_model <- df_trials_model %>%
  mutate(z_target_pointed_distance_difference = abs(scale(target_pointed_distance_difference)[,1])) %>%
  filter(z_target_pointed_distance_difference < 3)

lmer_pointed_distance <- lmer(
  target_pointed_distance_difference ~ OpticFlow + LevelSize + Vista +
    OpticFlow:LevelSize + OpticFlow:Vista + scale(pointingpoint_target_distance) +
    (1|participant),
  data = df_distance_model
)

distance_model_table <- summary(report_table(lmer_pointed_distance))
```

```{r pointed-distance-boxplot}
#| fig-cap: "Pointing distance errors by experimental conditions. Positive values indicate overshooting, negative values indicate undershooting."
df_distance_model %>%
  ggplot(aes(x = LevelSize, y = target_pointed_distance_difference, fill = OpticFlow, color = OpticFlow)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~Vista) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Pointing Distance Error by Condition",
    y = "Distance Error (meters)",
    x = "Environment Size",
    fill = "Movement Mode",
    color = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```


```{r pointed-distance-model-check}
#| fig-cap: "Model diagnostics for the linear mixed-effects model of pointing distance error."
#| fig-height: 12
#| fig-width: 8
check_model(lmer_pointed_distance)
```

```{r lmer-pointed-distance-table}
#| tbl-cap: "Linear mixed-effects model results for pointing distance error" 
report_model_parameters(distance_model_table) %>%
  tab_header(title = "Fixed-Effect Predictor Estimates for Pointing Distance Error Model")
report_model_summary_parameters(distance_model_table) %>%
  tab_header(title = "Model Parameters for Pointing Distance Error")
```

### Estimated Marginal Means

```{r lmer-pointed-distance-predictions}
#| tbl-cap: "Predicted distance errors across all experimental conditions"
predict_my_responses(lmer_pointed_distance) %>%
  report_my_predictions() %>%
  gt() %>%
    tab_header(title = "Predicted Distance Errors Across Experimental Conditions")

report_group_predictions(lmer_pointed_distance) %>% 
  group_by(Predictor) %>%
  gt() %>%
    tab_header(title = "Predicted Effects of Experimental Conditions on Pointing Distance Error")
```

### Key Findings

Participants made smaller distance estimation errors in vista spaces (environments with open sightlines) compared to non-vista spaces, regardless of movement mode or environment size. This suggests that visual access to the environment improves distance estimation accuracy.

## Pointing Angle Error

Next, we examined the angular accuracy of participants' pointing responses. Angle errors were computed as the absolute difference between the correct direction and the direction indicated by participants.

```{r angle-error-histogram}
#| fig-cap: "Distribution of absolute angle errors (log-transformed). The distribution shows positive skew, justifying the use of a Gamma regression model."
ggplot(df_trials_model) +
  geom_histogram(aes(x = log(abs(pointed_angle_difference))), bins = 30, fill = "coral", alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Absolute Angle Errors (Log Scale)",
    x = "Log(Angle Error in degrees)",
    y = "Frequency"
  )
```

```{r angle-error-z-scaled}
ggplot(df_trials_model, aes(pointed_angle_difference, fill = abs(scale(pointed_angle_difference)) > 3)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Raw Angle Errors",
    x = "Angle Error (degrees)",
    y = "Frequency"
  )
```

### Statistical Model

Due to the skewed distribution of the absolute angle errors, we used a generalized linear mixed-effects model with a Gamma distribution (as the log of the errors are left skewed) and a log link.

As the original angle errors were approximately symmetrically distributed around zero, we removed outliers based on the raw data. 

Extreme outliers (|z| > 3 away from the mean using the angle estimated) were excluded from the analysis.

```{r glmer-pointed-angle, results='hide'}
# Fit the full model with Gamma family
df_trials_model_angle <- df_trials_model %>%
  mutate(z_abs_pointed_angle_difference = abs(scale(pointed_angle_difference)),
        abs_pointed_angle_difference = abs(pointed_angle_difference)) %>%
  filter(z_abs_pointed_angle_difference < 3)

lmer_pointed_angle <- df_trials_model_angle %>%
  glmer(
    abs_pointed_angle_difference ~ OpticFlow + LevelSize + Vista +
      OpticFlow:LevelSize + OpticFlow:Vista + (1|participant),
    data = .,
    family = Gamma(link = "log")
  )
pointed_angle_model_table <- summary(report_table(lmer_pointed_angle))
```

```{r pointed-angle-boxplot}
#| fig-cap: "Absolute pointing angle errors by experimental conditions after outlier exclusion used in the model."
df_trials_model_angle %>%
  ggplot(aes(x = LevelSize, y = log(abs_pointed_angle_difference), fill = OpticFlow, color = OpticFlow)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~Vista) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Absolute Pointing Angle Error (Log) by Condition",
    y = "Log(Absolute Angle Error in degrees)",
    x = "Environment Size",
    fill = "Movement Mode",
    color = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```

```{r glmer-pointed-angle-check}
#| fig-cap: "Model diagnostics for the generalized linear mixed-effects model of pointing angle error (Gamma family, log link)."
#| fig-height: 12
#| fig-width: 8  
check_model(lmer_pointed_angle)
```

```{r glmer-pointed-angle-predictor-table}
#| tbl-cap: "Fixed-effect predictor estimates for pointing angle error model"
report_model_parameters(pointed_angle_model_table) %>%
  tab_header(title = "Fixed-Effect Predictor Estimates for Pointing Angle Error Model")
```

```{r glmer-pointed-angle-model-fit-table}
#| tbl-cap: "Model-level parameters and fit indices for pointing angle error model"
report_model_summary_parameters(pointed_angle_model_table) %>%
  tab_header(title = "Model Parameters for Pointing Angle Error")
```

### Estimated Marginal Means

```{r lmer-pointed-angle-predictions}
#| tbl-cap: "Predicted angle errors (in degrees) across experimental conditions"
predict_my_responses(lmer_pointed_angle) %>%
  report_my_predictions() %>%
  gt() %>%
    tab_header(title = "Predicted Angle Errors Across Experimental Conditions")
```

```{r angle-predictions-individual}
#| tbl-cap: "Marginal predictions for individual factors"
report_group_predictions(lmer_pointed_angle) %>% 
  group_by(Predictor) %>%
  gt() %>%
    tab_header(title = "Predicted Effects of Experimental Conditions on Pointing Angle Error")
```

### Key Findings

Angle errors were significantly affected by environment size, with larger environments leading to larger angular errors. However, neither movement mode nor vista condition significantly affected angular accuracy. This suggests that angular pointing is more sensitive to spatial scale than to visual access or movement information.

## Pointing Response Time

We examined how long participants took to make their pointing responses, which may reflect the cognitive difficulty of the spatial judgment.

```{r pointing-duration-analysis, results='hide'}
#| fig-cap: "Distribution of pointing response times (log-transformed) by movement mode"
ggplot(df_trials_model, aes(x = log(pointing_duration), fill = OpticFlow)) +
  geom_histogram(position="identity", bins = 30, alpha = 0.5) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Pointing Response Time Distribution",
    x = "Log(Response Time in seconds)",
    y = "Frequency",
    fill = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set2")
```

### Statistical Model

Given the left-skewed distribution of response times, we used a generalized linear mixed-effects model with a Gamma distribution and log link to analyze pointing durations. The data were not originally normally distributed, so we used outlier removel based on z-scores on the log transformed data. We excluded extreme outliers (|z| > 3) from the analysis.

```{r lmer-pointing-duration, results='hide'}
df_trials_model_point_duration <- df_trials_model %>%
  mutate(z_log_pointing_duration = abs(scale(log(pointing_duration)))) %>%
  filter(z_log_pointing_duration < 3)

lmer_point_duration <- glmer(
  pointing_duration ~ OpticFlow + LevelSize + Vista + OpticFlow:LevelSize + OpticFlow:Vista + (1|participant),
  data = df_trials_model_point_duration,
  #data = filter(df_trials_model, run != "Run3"),  # exclude participant with extreme outliers
  family = Gamma(link = "log")
)
pointing_duration_model_table <- summary(report_table(lmer_point_duration))
```

```{r pointing-duration-boxplot}
#| fig-cap: "Pointing response times by experimental conditions after outlier exclusion used in the model."
df_trials_model_point_duration %>%
  ggplot(aes(x = LevelSize, y = log(pointing_duration), fill = OpticFlow, color = OpticFlow)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~Vista) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Pointing Response Time (Log) by Condition",
    y = "Log(Response Time in seconds)",
    x = "Environment Size",
    fill = "Movement Mode",
    color = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```

```{r lmer-pointing-duration-check}
#| fig-cap: "Model diagnostics for the generalized linear mixed-effects model of pointing response time (Gamma family, log link)."
#| fig-height: 12
#| fig-width: 8
check_model(lmer_point_duration)
```

```{r lmer-pointing-duration-predictor-table}
#| tbl-cap: "Fixed-effect predictor estimates for pointing response time model"
report_model_parameters(pointing_duration_model_table) %>%
  tab_header(title = "Fixed-Effect Predictor Estimates for Pointing Response Time Model")
```

```{r lmer-pointing-duration-model-fit-table}
#| tbl-cap: "Model-level parameters and fit indices for pointing response time model"
report_model_summary_parameters(pointing_duration_model_table) %>%
  tab_header(title = "Model Parameters for Pointing Response Time")
```

### Estimated Marginal Means

```{r lmer-pointing-duration-predictions}
#| tbl-cap: "Predicted pointing response times (in seconds) across experimental conditions"
predict_my_responses(lmer_point_duration) %>%
  report_my_predictions() %>%
  gt() %>%
    tab_header(title = "Predicted Pointing Response Times Across Experimental Conditions")
```

```{r point-duration-predictions-individual}
#| tbl-cap: "Marginal predictions for individual factors"
report_group_predictions(lmer_point_duration) %>%
  group_by(Predictor) %>%
  gt() %>%
    tab_header(title = "Predicted Effects of Experimental Conditions on Pointing Response Time")
```

### Key Findings
Response times showed considerable variability across conditions, but no clear systematic patterns emerged from the model. The pointing was significantly slowere following large levels and following No optic flow movement. 

## Navigation Distance

We analyzed the total distance participants traveled during the search and return phases of each trial.

```{r navigated-distance-analysis, results='hide'}
#| fig-cap: "Distribution of total navigation distance across trials"
df_data %>%
  mutate(total_distance = distance_items + distance_return) %>%
  ggplot() +
  geom_histogram(aes(x = log(total_distance), fill = OpticFlow), bins = 30, alpha = 0.5, position = "identity") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  labs(
    title = "Distribution of Log of Navigated Distance",
    x = "Log(Distance Traveled in meters)",
    y = "Frequency",
    fill = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1")
```

### Statistical Model

We used a mixed effect linear model with a Gamma distribution and log link to analyze the total navigation distance. 

To remove extreme outliers, we excluded trials where the z-scored log-transformed total distance was greater than 3.

```{r glm-distance-analysis, results='hide'}
# Fit GLM with Gamma family
df_navdistance_model <- df_data %>%
  mutate(total_distance = (distance_items + distance_return),
         z_log_total_distance = abs(scale(log(total_distance)))) %>%
  filter(z_log_total_distance < 3)

lmer_traveled_distance <- glmer(total_distance ~ OpticFlow + Vista + LevelSize +
    OpticFlow:Vista + OpticFlow:LevelSize + (1|participant),
  data = df_navdistance_model,
  family = Gamma(link = "log"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

navdistance_model_table <- summary(report_table(lmer_traveled_distance))
```

```{r navigation-distance-boxplot}
#| fig-cap: "Navigation distance by experimental conditions after outlier exclusion used in the model."
df_navdistance_model %>%
  ggplot(aes(x = LevelSize, y = log(total_distance), fill = OpticFlow, color = OpticFlow)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~Vista) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  labs(
    title = "Navigation Distance (Log) by Condition",
    y = "Log(Distance Traveled in meters)",
    x = "Environment Size",
    fill = "Movement Mode",
    color = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```

```{r glm-distance-check}
#| fig-cap: "Model diagnostics for the generalized linear model of navigation distance (Gamma family, log link)."
#| fig-height: 12
#| fig-width: 8
check_model(lmer_traveled_distance)
```

```{r glm-distance-model-fit-table}
#| tbl-cap: "Model-level parameters and fit indices for navigation distance model"
report_model_summary_parameters(navdistance_model_table) %>%
  tab_header(title = "Model Parameters for Navigation Distance")
```

```{r glm-distance-predictor-table}
#| tbl-cap: "Fixed-effect predictor estimates for navigation distance model"
report_model_parameters(navdistance_model_table) %>%
  tab_header(title = "Fixed-Effect Predictor Estimates for Navigation Distance Model")
```

### Estimated Marginal Means

```{r glm-distance-predictions}
#| tbl-cap: "Predicted navigation distances across conditions using GLM with Gamma family and log link"
predict_my_responses(lmer_traveled_distance) %>%
  report_my_predictions() %>%
  gt() %>%
    tab_header(title = "Predicted Navigation Distances Across Experimental Conditions")

```

```{r navigation-distance-predictions-individual}
#| tbl-cap: "Marginal predictions for individual factors"
report_group_predictions(lmer_traveled_distance) %>%
  group_by(Predictor) %>%
  gt() %>%
    tab_header(title = "Predicted Effects of Experimental Conditions on Navigation Distance")
```

### Key Findings
The presence of optic flow did not significantly affect the total distance traveled. However, participants traveled significantly shorter distances in vista spaces compared to non-vista spaces, suggesting more efficient navigation when environmental layout is visible.

## Number of Teleports

We also examined the number of teleports participants used during navigation, which may reflect their strategy and confidence in navigating the environment.

```{r teleports-descriptive}
#| fig-cap: "Distribution of number of teleports across trials"
ggplot(df_data) +
  geom_histogram(aes(x = n_teleports, fill = OpticFlow), binwidth = 1, alpha = 0.7, position = "identity") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  labs(
    title = "Distribution of Number of Teleports",
    x = "Number of Teleports",
    y = "Frequency",
    fill = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  geom_vline(xintercept = mean(df_data$n_teleports, na.rm = TRUE) + 3 * sd(df_data$n_teleports, na.rm = TRUE),
             linetype = "dashed", color = "red")
```

Removing extreme outliers (|z|) > 3. As the number of teleports represents count data, we are using poisson distribution

### Statistical Model

We modeled the number of teleports using a mixed-effects generalized linear model with a Poisson distribution. 

```{r teleports-model, results='hide'}
df_model_teleports <- df_data %>%
  mutate(z_n_teleports = abs(scale(n_teleports)[,1])) %>%
  filter(z_n_teleports < 3)

lmer_teleports <- glmer(
  n_teleports ~ OpticFlow + Vista + LevelSize +
    OpticFlow:Vista + OpticFlow:LevelSize + (1|participant),
  data = df_model_teleports,
  family = poisson() 
)
check_overdispersion(lmer_teleports)
```

:::{.callout-warning}
Due to overdispersion in the Poisson model, we could refitted the model using a negative binomial distribution to account for the extra variability in the count data.
I am thinking of not doing this for clarity, but we very well might. It is as easilly interpretable in the predicted values
:::


We refitted the model using a negative binomial distribution to account for overdispersion.

```{r teleports-overdispersion-model, results='hide'}
lmer_teleports <- lme4::glmer.nb(
  n_teleports ~ OpticFlow + Vista + LevelSize +
    OpticFlow:Vista + OpticFlow:LevelSize + (1|participant),
  data = df_model_teleports
)

teleports_model_table <- summary(report_table(lmer_teleports))
# need to column z to T values so that the report_model_parameters function can work with it
teleports_model_table$t <- teleports_model_table$z
```

```{r teleports-boxplot}
#| fig-cap: "Number of teleports by experimental conditions after outlier exclusion used in the model."
df_model_teleports %>%
  ggplot(aes(x = LevelSize, y = n_teleports, fill = OpticFlow, color = OpticFlow)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~Vista) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  labs(
    title = "Number of Teleports by Condition",
    y = "Number of Teleports",
    x = "Environment Size",
    fill = "Movement Mode",
    color = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```

```{r glm-teleports-check}
#| fig-cap: "Model diagnostics for the generalized linear model of number of teleports (negative binomial)."
#| fig-height: 12
#| fig-width: 8
check_model(lmer_teleports)
```

```{r glm-teleports-model-fit-table}
#| tbl-cap: "Model-level parameters and fit indices for teleports model"
report_model_summary_parameters(teleports_model_table) %>%
  tab_header(title = "Model Parameters for Number of Teleports")
```

```{r glm-teleports-predictor-table}
#| tbl-cap: "Fixed-effect predictor estimates for teleports model"
report_model_parameters(teleports_model_table) %>%
  tab_header(title = "Fixed-Effect Predictor Estimates for Number of Teleports Model")
```

### Estimated Marginal Means

```{r glm-teleports-predictions}
#| tbl-cap: "Predicted number of teleports across conditions using negative binomial model"
predict_my_responses(lmer_teleports) %>%
  report_my_predictions() %>%
  gt() %>%
    tab_header(title = "Predicted Number of Teleports Across Experimental Conditions")
```

```{r teleports-predictions-individual}
#| tbl-cap: "Marginal predictions for individual factors"
report_group_predictions(lmer_teleports) %>%
  group_by(Predictor) %>%
  gt() %>%
    tab_header(title = "Predicted Effects of Experimental Conditions on Number of Teleports")
```

## Navigation Duration

Finally, we examined the total time participants spent navigating during the search and return phases.

```{r search-duration-histogram}
#| fig-cap: "Distribution of total navigation duration (log-transformed) by movement mode"
df_data %>%
  mutate(total_duration = duration_return + duration_searching) %>%
  ggplot(aes(x = log(total_duration), fill = OpticFlow)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  labs(
    title = "Total Navigation Duration Distribution",
    x = "Log(Duration in seconds)",
    y = "Frequency",
    fill = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1")
```

### Statistical Model

As the navigation duration data were also left-skewed and not normally distributed, we used a generalized mixed effect model with a Gamma distribution and log link. 

We excluded extreme outliers (|z| > 3 on the log-transformed total duration) from the analysis.

```{r glm-duration-analysis, results='hide'}
df_model_duration <- df_data %>%
  mutate(total_duration = duration_return + duration_searching,
         z_log_total_duration = abs(scale(log(total_duration)))) %>%
  filter(z_log_total_duration < 3)

lmer_navigated_duration <- glmer(
  total_duration ~ OpticFlow + Vista + LevelSize +
    OpticFlow:Vista + OpticFlow:LevelSize + (1|participant),
  data = df_model_duration,
  family = Gamma(link = "log")
)

duration_model_table <- summary(report_table(lmer_navigated_duration))
```

```{r navigation-duration-boxplot}
#| fig-cap: "Navigation duration by experimental conditions after outlier exclusion used in the model."
df_model_duration %>%
  ggplot(aes(x = LevelSize, y = log(total_duration), fill = OpticFlow, color = OpticFlow)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3) +
  facet_wrap(~Vista) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom") +
  labs(
    title = "Navigation Duration (Log) by Condition",
    y = "Log(Duration in seconds)",
    x = "Environment Size",
    fill = "Movement Mode",
    color = "Movement Mode"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1")
```

```{r glm-duration-check}
#| fig-cap: "Model diagnostics for the generalized linear model of navigation duration (Gamma family, log link)."
#| fig-height: 12
#| fig-width: 8
check_model(lmer_navigated_duration)
```

```{r glm-duration-model-fit-table}
#| tbl-cap: "Model-level parameters and fit indices for navigation duration model"
report_model_summary_parameters(duration_model_table) %>%
  tab_header(title = "Model Parameters for Navigation Duration")
```

```{r glm-duration-predictor-table}
#| tbl-cap: "Fixed-effect predictor estimates for navigation duration model"
report_model_parameters(duration_model_table) %>%
  tab_header(title = "Fixed-Effect Predictor Estimates for Navigation Duration Model")
```

### Estimated Marginal Means

```{r glm-duration-predictions}
#| tbl-cap: "Predicted navigation durations across conditions"
predict_my_responses(lmer_navigated_duration) %>%
  report_my_predictions() %>%
  gt() %>%
    tab_header(title = "Predicted Navigation Durations Across Experimental Conditions")
```

```{r navigation-duration-predictions-individual}
#| tbl-cap: "Marginal predictions for individual factors"
report_group_predictions(lmer_navigated_duration) %>%
  group_by(Predictor) %>%
  gt() %>%
    tab_header(title = "Predicted Effects of Experimental Conditions on Navigation Duration")
```

### Key Findings

Only the level size was a significant predictor of navigation time, unlike in the results of navigated distance, where also vista condition was significant. This suggests that while larger environments naturally require more time to navigate, the visibility of the environment does not substantially impact how long participants take to complete navigation. This overall suggests that participants may adjust their speed based on environmental complexity and potential move faster in non-vista spaces.

## Simulator Sickness

We assessed simulator sickness using the SSQ and VRSQ questionnaires to examine potential adverse effects of the different movement modes.

```{r ssq-descriptives}
#| tbl-cap: Simulator sickness scores by movement mode (SSQ metrics)
df_questionnaires %>%
  select(starts_with("ssq_"), starts_with("vrsq_"), Movement) %>%
  pivot_longer(cols = -Movement, names_to = "ssq_metric", values_to = "value") %>%
  group_by(ssq_metric, Movement) %>%
  summarize(
    Mean = round(mean(value, na.rm = TRUE), ROUNDING_DIGITS),
    SD = round(sd(value, na.rm = TRUE), ROUNDING_DIGITS),
    mean_sd = paste0(Mean, " (", SD, ")"),
    .groups = "drop"
  ) %>%
  select(-c(Mean, SD)) %>%
  pivot_wider(names_from = Movement, values_from = mean_sd) %>%
  select(ssq_metric, Teleport, OpticFlow) %>%
  rename(
    `SSQ Metric` = ssq_metric,
    `Without Optic Flow` = Teleport,
    `With Optic Flow` = OpticFlow,
  ) %>%
  gt() %>%
    tab_header(title = "Simulator Sickness Scores by Movement Mode (SSQ and VRSQ)")
```

### Statistical Tests

We used simple paired t-tests to compare simulator sickness scores between the two movement modes for each SSQ and VRSQ metric.

```{r ssq-tests}
#| tbl-cap: "Paired t-tests comparing simulator sickness between movement modes"
df_questionnaires %>%
  select(c(starts_with("ssq_"), starts_with("vrsq_")), participant, Movement) %>%
  pivot_longer(cols = -c(participant, Movement), names_to = "ssq_metric", values_to = "value") %>%
  pivot_wider(names_from = Movement, values_from = value, id_cols = c(participant, ssq_metric)) %>%
  group_by(ssq_metric) %>%
  nest() %>%
  mutate(
    test = map(data, ~ t.test(.x$OpticFlow, .x$Teleport, paired = TRUE) %>% tidy())
  ) %>%
  select(ssq_metric, test) %>%
  unnest(test) %>%
  select(
    Metric = ssq_metric,
    `Mean Difference` = estimate,
    `t-statistic` = statistic,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  ) %>%
  mutate(
    `Mean Difference` = round(`Mean Difference`, ROUNDING_DIGITS),
    `t-statistic` = round(`t-statistic`, ROUNDING_DIGITS),
    `p-value` = round(`p-value`, ROUNDING_DIGITS),
    `95% CI Lower` = round(`95% CI Lower`, ROUNDING_DIGITS),
    `95% CI Upper` = round(`95% CI Upper`, ROUNDING_DIGITS)
  ) %>%

  knitr::kable(align = "lccccc")
```

Looking at the results from the paired t-tests, there were multiple significant differences in simulator sickness scores between the two movement modes across all SSQ and VRSQ metrics. 

While these reulst would not survive comparison for multiple corrections, we can safely asy that the Added optic flow to the teleportation movement mode increased simulator sickness symptoms in participants except the oculomotor strain subscale of the SSQ and VRSQ.

## VR Embodiment and Presence (VRLEQ)

```{r vrleq-descriptives}
#| tbl-cap: "VR embodiment and presence scores by movement mode"
df_questionnaires %>%
  select(starts_with("vrleq_"), Movement, participant) %>%
  pivot_longer(cols = -c(Movement, participant), names_to = "vrleq_metric", values_to = "value") %>%
  group_by(vrleq_metric, Movement) %>%
  summarize(
    Mean = round(mean(value, na.rm = TRUE), ROUNDING_DIGITS),
    SD = round(sd(value, na.rm = TRUE), ROUNDING_DIGITS),
    mean_sd = paste0(Mean, " (", SD, ")"),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    Range = paste0("(", Min, ", ", Max, ")"),
    .groups = "drop"
  ) %>%
  select(-c(Mean, SD, Min, Max)) %>%
  pivot_wider(names_from = Movement, values_from = c(mean_sd, Range)) %>%
  select(
    vrleq_metric,
    mean_sd_Teleport, Range_Teleport,
    mean_sd_OpticFlow, Range_OpticFlow
  ) %>%
  rename(
    `VRLEQ Metric` = vrleq_metric,
    `Mean (Withouth OF)` = mean_sd_Teleport,
    `Range (Withouth OF)` = Range_Teleport,
    `Mean (With OF)` = mean_sd_OpticFlow,
    `Range (With OF)` = Range_OpticFlow
  ) %>%
  gt() %>%
    tab_header(title = "VR Embodiment and Presence Scores by Movement Mode (VRLEQ)")
```

```{r vrleq-tests}
#| tbl-cap: "Paired t-tests comparing VR embodiment and presence between movement modes"
df_questionnaires %>%
  select(starts_with("vrleq_"), participant, Movement) %>%
  pivot_longer(cols = -c(participant, Movement), names_to = "vrleq_metric", values_to = "value") %>%
  pivot_wider(names_from = Movement, values_from = value, id_cols = c(participant, vrleq_metric)) %>%
  group_by(vrleq_metric) %>%
  nest() %>%
  mutate(
    test = map(data, ~ t.test(.x$OpticFlow, .x$Teleport, paired = TRUE) %>% tidy())
  ) %>%
  select(vrleq_metric, test) %>%
  unnest(test) %>%
  select(
    Metric = vrleq_metric,
    `Mean Difference` = estimate,
    `t-statistic` = statistic,
    `p-value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  ) %>%
  mutate(
    `Mean Difference` = round(`Mean Difference`, ROUNDING_DIGITS),
    `t-statistic` = round(`t-statistic`, ROUNDING_DIGITS),
    `p-value` = round(`p-value`, ROUNDING_DIGITS),
    `95% CI Lower` = round(`95% CI Lower`, ROUNDING_DIGITS),
    `95% CI Upper` = round(`95% CI Upper`, ROUNDING_DIGITS)
  ) %>%
  gt() %>%
    tab_header(title = "Paired t-tests Comparing VR Embodiment and Presence Between Movement Modes (VRLEQ)")
```

# Other Analyses

## Gender Differences

### Outcome measures differences

```{r gender-all-metrics}
#| fig-cap: "Distribution of spatial performance metrics by gender "
#| fig-height: 8
#| fig-width: 8
df_gender_comparisons <- df_trials_model %>%
  select(participant, Gender, 
         target_pointed_distance_difference,
         log_abs_pointed_angle_difference = pointed_angle_difference,
         pointing_duration) %>%
  mutate(log_abs_pointed_angle_difference = log(abs(log_abs_pointed_angle_difference))) %>%
  pivot_longer(cols = c(target_pointed_distance_difference,
                        log_abs_pointed_angle_difference,
                        pointing_duration),
               names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  mutate(abs_z_value = abs(scale(value))) %>%
  ungroup() %>%
  filter(abs_z_value < 3)

ggplot(df_gender_comparisons, aes(x = value, fill = Gender)) +
  geom_histogram(aes(y=after_stat(density)), position = "identity", bins = 30, alpha = 0.7) +
  facet_wrap(~metric, scales = "free", ncol = 1) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Spatial Performance Metrics by Gender",
    x = "Value",
    y = "Density"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")
```

There seem to be no significant difference between the genders, with the exception of pointing duration where males seem to bave slightly slower response times.

### Simulator Sickness by Gender

```{r ssq-gender}
#| tbl-cap: "Simulator sickness scores by gender (SSQ metrics)"
df_questionnaires %>%
  select(starts_with("vrsq_"), Gender, OpticFlow) %>%
  pivot_longer(cols = -c(Gender, OpticFlow), names_to = "ssq_metric", values_to = "value") %>%
  ggplot(aes(y = value, fill = Gender)) +
    geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
    facet_wrap(~ssq_metric, scales = "free") +
    theme_minimal(base_size = 14) +
    labs(title = "Simulator Sickness Scores by Gender and Movement Mode") + 
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")
```

### Gender and optic flow interaction

```{r gender-opticflow-interaction, fig.height=10}
#| fig-cap: "Simulator sickness scores by gender and movement mode"
df_questionnaires %>%
  select(starts_with("vrsq_"), Gender, OpticFlow) %>%
  pivot_longer(cols = -c(Gender, OpticFlow), names_to = "ssq_metric", values_to = "value") %>%
  ggplot(aes(x = OpticFlow,y = value, fill = OpticFlow)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  facet_wrap(~ssq_metric + Gender, scales = "free", ncol = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Simulator Sickness Scores by Gender and Movement Mode",
    x = "Score",
    y = "Frequency"
  ) + 
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

There does not seem to be any strong gender pattern in the simulator sickness scores, nor any strong interaction between